---
title: "Air Quality Analysis"
author: "Ryan Nathaniel"
date: "2026-03-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dataset Overview

```{r cars}
# Load ggplot2 package
library(ggplot2)

getwd()
setwd('/Users/ryannathaniel/NTU/basic_scRNAseq/Week2/Assignment')

# Read the data and store it in df
df <- read.csv("airquality.csv")

# To see the number of records
nrow(df)

# To see all the variables exist in the data
names(df)
```

## Barplot
This barplot helps visualize how the average environmental conditions differ by month.

```{r}
# Average calculation for each varible
mean_ozone <- tapply(df$Ozone, df$Month, mean, na.rm = TRUE)
mean_solar <- tapply(df$Solar.R, df$Month, mean, na.rm = TRUE)
mean_wind <- tapply(df$Wind, df$Month, mean, na.rm = TRUE)
mean_temp <- tapply(df$Temp, df$Month, mean, na.rm = TRUE)

# Create a new data frame for the barplot
plot_data <- data.frame(
  month = factor(rep(names(mean_ozone), 4)),
  var = c(rep("Ozone", 5),
          rep("Solar.R", 5),
          rep("Wind", 5),
          rep("Temp", 5)),
  val = c(mean_ozone, mean_solar, mean_wind, mean_temp)
)

# Create the barplot
ggplot(plot_data, aes(x = month, y = val, fill = var)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Average Value of Each Variable by Month",
       x = "Month",
       y = "Average Value",
       fill = "Variable") +
  theme_minimal()
```

## Boxplot
This boxplot helps visualize the distribution of ozone levels in each month, including the median, spread, and potential outliers.

```{r}
# Create the boxplot
ggplot(df, aes(x = factor(Month), y = Ozone)) +
  geom_boxplot(fill = "white", na.rm = TRUE) +
  geom_hline(yintercept = 54, linetype = "dashed", color = "red") +
  annotate("text", x = 5, y = 58, label = "Reference = 54", color = "red") +
  labs(title = "Distribution of Ozone by Month",
       x = "Month",
       y = "Ozone") +
  theme_minimal()
```

## PCA
This PCA plot shows that observations from different months overlap substantially, suggesting that the monthly groups are not completely separated by these variables.

```{r}
# Select the variables for PCA
pca_data <- df[, c("Ozone", "Solar.R", "Wind", "Temp")]

# Keep only complete rows
complete_rows <- complete.cases(pca_data)
pca_data <- pca_data[complete_rows, ]

# Run PCA
pca_result <- prcomp(pca_data, scale. = TRUE)
summary(pca_result)

# Create a data frame for plotting
pca_plot_data <- data.frame(
  PC1 = pca_result$x[, 1],
  PC2 = pca_result$x[, 2],
  Month = factor(df$Month[complete_rows])
)

# Create PCA plot
ggplot(pca_plot_data, aes(x = PC1, y = PC2, color = Month)) +
  geom_point(size = 3) +
  labs(title = "PCA Plot for Air Quality",
       x = "PC1",
       y = "PC2",
       color = "Month") +
  theme_minimal()
```

## ComplexHeatmap
This ComplexHeatmap helps summarize the monthly patterns of ozone, solar radiation, wind, and temperature by showing which variables are relatively high or low in each month.

```{r}
# Load packages
library(ComplexHeatmap)
library(circlize)

# Combine into a data frame
heatmap_data <- data.frame(
  Ozone = mean_ozone,
  Solar.R = mean_solar,
  Wind = mean_wind,
  Temp = mean_temp
)

# Sort rows by month
heatmap_data <- heatmap_data[order(as.numeric(rownames(heatmap_data))), ]

# Convert to matrix
mat <- as.matrix(heatmap_data)

# Scale the values (Z-score)
mat_scaled <- scale(mat)

# Set row names
rownames(mat_scaled) <- paste0("Month ", rownames(heatmap_data))

# Create right annotation for monthly average ozone
right_ha <- rowAnnotation(
  AvgOzone = anno_barplot(as.numeric(heatmap_data$Ozone),
                          gp = gpar(fill = "gray70", col = "black"),
                          border = FALSE)
)

# Create heatmap
Heatmap(mat_scaled,
        name = "Z-score",
        right_annotation = right_ha,
        col = colorRamp2(c(-2, 0, 2), c("blue", "white", "red")),
        cluster_rows = FALSE,
        cluster_columns = TRUE,
        row_title = "Month",
        column_title = "Monthly Average of Air Quality Variables")
```

